---
title: "pset1"
author: "Chen Dong"
date: "10/25/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ANSWER 1.A.1
$3/6=0.5$

# ANSWER 1.A.2
$(1/6)*(2/6)+(1/6)*(1/6)=1/12$

# ANSWER 1.B.1
Binomial distribution. X is discrete variable.

# ANSWER 1.B.2
$Pr(X = k) =  \frac{n!}{k!(n-k)!}{p^k}(1-p)^{n-k} ~~for ~~(0 \leq k \leq n)$

# CODE 1.B.3
```{r}
dbinom(4,20,0.2)
pbinom(4,20,0.2,lower.tail = F)
```

# ANSWER 1.B.3
$Pr(X = 4)=0.2181994$ and $Pr(X \geq 4)=0.3703517$

# CODE 2.A.1
```{r}
par(mfrow=c(1,2))
plot(function(x) dnorm(x,2.5,sqrt(2.25)), -2,7,main = "Density curve", ylab = "")
plot(function(x) pnorm(x,2.5,sqrt(2.25)), -2,7,main = "Cumulative probability curve", ylab = "")
```

# CODE 2.A.2
```{r}
pnorm(4,2.5,sqrt(2.25))-pnorm(0.5,2.5,sqrt(2.25))
```

# CODE 2.A.3
```{r}
pnorm(log(30),2.5,sqrt(2.25),lower.tail = F)
```

# CODE 2.A.4
```{r}
qnorm(0.9,2.5,sqrt(2.25))
```

# CODE 2.B.1
```{r}
pnorm(80.5,90,38,lower.tail = F)
```
# CODE 2.B.2
```{r}
2*pnorm(90-1.5*38,90,38)
```

# CODE 2.B.3
```{r}
(90-qnorm(0.04/2,90,38))/38
```

# CODE 3.1.1
```{r}
m0 <- rnorm(1000,0,1)
m10 <- numeric()
m100 <- numeric()
m1000 <- numeric()
for (i in 1:1000){
  m10[i] <- mean(rnorm(10,0,1))
  m100[i] <- mean(rnorm(100,0,1))
  m1000[i] <- mean(rnorm(1000,0,1))
}
par(mfrow=c(2,2))
hist(m0)
hist(m10)
hist(m100)
hist(m1000)
```

# CODE 3.1.2
```{r}
qqnorm(m0)
```

# ANSWER 3.1.2
The distribution looks normal.

# CODE 3.1.3
```{r}
m <- data.frame(m0,m10,m100,m1000)
# mean
mmean <- apply(m, 2, mean)
mmean
# var
mvar <- apply(m, 2, var)
mvar
# var ratio
mvar[1]/mvar[2:4]
```
# ANSWER 3.1.3
The means are not substantially different from each other.
The variances are different from each other. 
The ratio between Var(m0) and Var(m10),Var(m100),Var(m1000) are about 10,100,1000.
A general formula to derive the Variance of any distribution of the means X¯n for any given n would be $\frac{\sigma}{n}$

# CODE 3.2.1
```{r}
r0 <- rexp(1000,1)
r10 <- numeric()
r100 <- numeric()
r1000 <- numeric()
for (i in 1:1000){
  r10[i] <- mean(rexp(10,1))
  r100[i] <- mean(rexp(100,1))
  r1000[i] <- mean(rexp(1000,1))
}
par(mfrow=c(2,2))
hist(r0)
hist(r10)
hist(r100)
hist(r1000)
```

# CODE 3.2.2
```{r}
par(mfrow=c(2,2))
qqnorm(r0)
qqnorm(r10)
qqnorm(r100)
qqnorm(r1000)
```

# ANSWER 3.2.2
The exponential values are not normally distributed. The means are normally distributed.


# CODE 3.2.3
```{r}
r <- data.frame(r0,r10,r100,r1000)
# mean
rmean <- apply(r, 2, mean)
rmean
# var
rvar <- apply(r, 2, var)
rvar
# var ratio
rvar[1]/rvar[2:4]
```
# ANSWER 3.2.3
The means are not substantially different from each other.
The variances are different from each other. 
The ratio between Var(r0) and Var(r10),Var(r100),Var(r1000) are about 10,100,1000.
