---
title: 'Hypothesis testing: some extra examples'
author: "Eric Bartell"
output:
  html_document:
    code_folding: show
    fig_height: 8
    fig_width: 10
    highlight: haddock
    toc: yes
    toc_depth: 5
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
  word_document:
    toc: yes
    toc_depth: '5'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE , eval=TRUE)
```


## One Sample

We know the average height of an adult male in the US is 70 inches with a standard deviation of 3 inches.

If we want to check if the heights of people in our class is the same as that of the US average male, what statistical test should we use?

$$Z=\frac{\bar{X}-\mu}{\frac{\sigma}{\sqrt{n}}}$$
$$t=\frac{\bar{X}-\mu}{\frac{s}{\sqrt{n}}}$$

My plan here is to first look at how these function with a couple of class heights (2-3).

As we investigate, what do you notice about the difference in the p-values from the Z-test vs the t-test? Why is this?
```
The Z-test should generally be more significant than the t-test because they follow different distributions.
```
Then we will look at generating heights, and as n increases p-values will drop for both tests.

```{r, echo=T, eval=T}
set.seed(1)
mu=70
#heights=c(72,66,66.5,68)
#heights=rnorm(n = 2,mean = 64,sd = 3)
#heights=rnorm(n = 10,mean = 64,sd = 3)
heights=rnorm(n = 50,mean = 64,sd = 3)


xbar=mean(heights)
populationStandardDeviation=3
n=length(heights)

#Trying Z-test:
zscore=abs(xbar-mu)/(populationStandardDeviation/sqrt(n))
pval_z=1-pnorm(zscore)
print("Z-test pvalue:")
print(pval_z)

#Trying T-test:
sampleStandardDeviation=sd(heights)
tstat=abs(xbar-mu)/(sampleStandardDeviation/sqrt(n))
pval_t=1-pt(tstat,df = n-1)
print("T-test pvalue:")
print(pval_t)

#t.test(x = heights, mu = 70, alternative = "less")

```


## Two Sample

We are investigating the mutation rate of different cancer cell lines. We have, for each cell type, multiple counts of number of mutations identified in that cell line.

### 1
We prepare replicates of cell type A and B together, and we do this multiple times (so we have n pairs of cells of types A and B that were grown in parallel). In this case, what test would we use?

### 2.1
We prepare batches of A and then batches of B; we have an uneven number of replicates from each group. Which test do we use here?

Non-equal variances:
$$t=\frac{\bar{x_1}-\bar{x_2}}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}$$
where
$$df_{not equal}=\frac{(\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2})^2}{\frac{(\frac{s_1^2}{n_1})^2}{n_1-1} + \frac{(\frac{s_2^2}{n_2})^2}{n_2-1}}$$

Equal variances:
$$t=\frac{\bar{x_1}-\bar{x_2}}{s\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}}$$
where
$$s=\sqrt{\frac{(n_1-1)s_1^2 + (n_2-1)s_2^2}{n_1+n_2-2}}$$
and
$$df_{equal}=n_1 + n_2 - 2$$

Do we perform a test for equal variances or different variances?
$$F=\frac{s_1^2}{s_2^2}$$
where
$$df_1=n_1-1$$
$$df_2=n_2-2$$

### 2.2
Perform this test!
Are the A and B groups mutation counts significantly different?

```{r}
set.seed(2)
#(Cervical cancer: ~8.6 mutation per megabase * ~3000 megabases)
A_numMutations = rnorm(8,mean = 25800,sd = 5500) 

#(Esophagial cancer: ~9.2 mutation per megabase * ~3000 megabases)
B_numMutations = rnorm(7,mean = 27600,sd = 7200) 

x_1_bar=mean(A_numMutations)
x_2_bar=mean(B_numMutations)
n_1=length(A_numMutations)
n_2=length(B_numMutations)
s_1=sd(A_numMutations)
s_2=sd(B_numMutations)


#perform f_stat test
f_stat=s_1^2/s_2^2
f_pval=pf(f_stat,df1 = n_1-1, df2 = n_2-1)
print("F-stat p-value")
print(f_pval)

#for equal variances
s=sqrt(((n_1-1)*s_1^2 + (n_2-1)*s_2^2)/(n_1+n_2-2))

#... perform t-test
tstat=abs(x_1_bar-x_2_bar)/(s*sqrt(1/n_1 + 1/n_2))
t_pval=1-pt(tstat,df = n_1 + n_2 - 2)

print("T-test (equal variances) p-value")
print(t_pval)
```

### 2.3
How about if we increase our sample size?
```{r}
set.seed(2)
A_numMutations = rnorm(120,mean = 25800,sd = 5500)
B_numMutations = rnorm(113,mean = 27600,sd = 7200)
x_1_bar=mean(A_numMutations)
x_2_bar=mean(B_numMutations)
n_1=length(A_numMutations)
n_2=length(B_numMutations)
s_1=sd(A_numMutations)
s_2=sd(B_numMutations)

#perform f-stat test
f_stat=s_1^2/s_2^2
f_pval=pf(f_stat,df1 = n_1-1, df2 = n_2-1)
cat("F-stat p-value: ",f_pval,"\n")


# for unequal variances
df_unequal=(s_1^2/n_1 + s_2^2/n_2)^2/
  ((s_1^2/n_1)^2/(n_1-1) + (s_2^2/n_2)^2/(n_2-1))

#... perform t-test
tstat=abs(x_1_bar-x_2_bar)/(sqrt(s_1^2/n_1 + s_2^2/n_2))
t_pval=1-pt(tstat,df = df_unequal)

cat("T-test (unequal variances) p-value: ",t_pval, "\n")

#for equal variances
s=sqrt(((n_1-1)*s_1^2 + (n_2-1)*s_2^2)/(n_1+n_2-2))

#... perform t-test
tstat=abs(x_1_bar-x_2_bar)/(s*sqrt(1/n_1 + 1/n_2))
t_pval=1-pt(tstat,df = n_1 + n_2 - 2)

cat("T-test (equal variances) p-value: ", t_pval, "\n")

```

### 2.4
For our small sample size, what is our 95% confidence interval of the difference in mutation rate between the two populations? Does it include 0?
```{r}
set.seed(2)
A_numMutations = rnorm(8,mean = 25800,sd = 5500)
B_numMutations = rnorm(7,mean = 27600,sd = 7200)
x_1_bar=mean(A_numMutations)
x_2_bar=mean(B_numMutations)
n_1=length(A_numMutations)
n_2=length(B_numMutations)
s_1=sd(A_numMutations)
s_2=sd(B_numMutations)

#already performed f-stat: equal

#for unequal variances
s=sqrt(((n_1-1)*s_1^2 + (n_2-1)*s_2^2)/(n_1+n_2-2))

#... Confidence interval
delta_xbar = x_2_bar-x_1_bar
range = qt(.975,df = n_1 + n_2 - 2) * s * sqrt(1/n_1 + 1/n_2)
cat("We are 95% confident our true difference is between ", delta_xbar-range, " and ",delta_xbar+range, "\n")

```
### 2.5
How does our confidence interval change when we increase our sample size? 

```{r}
set.seed(2)
A_numMutations = rnorm(120,mean = 25800,sd = 5500)
B_numMutations = rnorm(113,mean = 27600,sd = 7200)
x_1_bar=mean(A_numMutations)
x_2_bar=mean(B_numMutations)
n_1=length(A_numMutations)
n_2=length(B_numMutations)
s_1=sd(A_numMutations)
s_2=sd(B_numMutations)

#already performed f-stat: equal

#for equal variances
df_unequal=(s_1^2/n_1 + s_2^2/n_2)^2/
  ((s_1^2/n_1)^2/(n_1-1) + (s_2^2/n_2)^2/(n_2-1))

#... Confidence interval
delta_xbar = x_2_bar-x_1_bar
range = qt(.975,df = df_unequal) * sqrt(s_1^2/n_1 + s_2^2/n_2)
cat("We are 95% confident our true difference is between ", delta_xbar-range, " and ",delta_xbar+range, "\n")

```

Cancer mutation rates inspired by Alexandrov et. al Nature 2013.